---
title: "Adaptive Importance Sampling for LOO-CV"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adaptive Importance Sampling for LOO-CV}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

The `bayesianquilts` R package provides **Adaptive Importance Sampling (AIS)**
methods for efficient Leave-One-Out cross-validation (LOO-CV).  When standard
PSIS-LOO produces high Pareto k-hat diagnostics (indicating unreliable
estimates), AIS applies a series of parameter-space transformations to improve
the importance weights.

The method is described in Chang et al. (2024),
[arXiv:2402.08151v2](https://arxiv.org/abs/2402.08151v2).

## Available Transformations

Transformations are tried in order of computational cost. Once a data point
achieves k-hat below the threshold (default 0.7), it is considered "adapted"
and subsequent transformations are skipped for that point.

| Name | Description | Cost |
|------|-------------|------|
| `identity` | Standard PSIS-LOO (baseline) | O(1) |
| `mm1` | Global moment matching (shift) | O(SN) |
| `mm2` | Global moment matching (shift + scale) | O(SN) |
| `pmm1` | Partial moment matching (shift) | O(SN) |
| `pmm2` | Partial moment matching (shift + scale) | O(SN) |
| `ll` | Likelihood descent (gradient-based) | O(SNK) |
| `kl` | KL divergence (posterior-weighted gradient) | O(SNK) |
| `nkl` | Natural-gradient KL (preconditioned by posterior covariance) | O(SNK + K^2) |

## Quick Start with brms

The simplest way to use AIS is with a `brms` model:

```{r brms-example, eval=FALSE}
library(bayesianquilts)
library(brms)

# Fit a logistic regression
fit <- brm(y ~ x1 + x2, data = mydata, family = bernoulli())

# Run AIS-LOO
results <- ais_brms(fit, verbose = TRUE)

# Best k-hat diagnostics
summary(results$best$khat)
```

## Detailed Example: Logistic Regression

```{r logistic-example}
library(R6)
library(loo)

# Source package files (or use library(bayesianquilts) if installed)
pkg_dir <- system.file(package = "bayesianquilts")
if (pkg_dir == "") {
  # Development: source files directly
  for (f in c("utils.R", "psis.R", "ais_classes.R", "transformations.R",
              "likelihoods.R", "ais.R")) {
    source(file.path("R", f))
  }
}

# Simulate data
set.seed(42)
N <- 100; K <- 3
beta_true <- c(1.5, -0.8, 0.5)
intercept_true <- 0.3

X <- matrix(rnorm(N * K), nrow = N)
eta <- X %*% beta_true + intercept_true
y <- rbinom(N, 1, 1 / (1 + exp(-eta)))
data <- list(X = X, y = y)

# Simulate posterior samples
S <- 200
beta_samples <- matrix(
  rnorm(S * K, mean = rep(beta_true, each = S), sd = 0.3),
  nrow = S, ncol = K
)
intercept_samples <- rnorm(S, mean = intercept_true, sd = 0.3)
params <- list(beta = beta_samples, intercept = intercept_samples)
```

### Create likelihood and run AIS

```{r run-ais}
lik <- LogisticRegressionLikelihood$new()
sampler <- AdaptiveImportanceSampler$new(lik)

results <- sampler$adaptive_is_loo(
  data, params,
  transformations = c("identity", "mm1", "ll", "kl", "nkl"),
  verbose = TRUE
)
```

### Inspect results

```{r results}
cat("Best k-hat: mean =", round(mean(results$best$khat), 3),
    ", max =", round(max(results$best$khat), 3), "\n")

# Compare transformations
for (nm in names(results)) {
  if (nm == "best" || is.null(results[[nm]]$khat)) next
  cat(sprintf("  %-22s  mean_khat=%.3f  max_khat=%.3f\n",
              nm, mean(results[[nm]]$khat), max(results[[nm]]$khat)))
}
```

## Custom Likelihood Functions

To use AIS with a custom model, implement the `LikelihoodFunction` interface:

```{r custom-likelihood, eval=FALSE}
MyLikelihood <- R6::R6Class("MyLikelihood",
  inherit = LikelihoodFunction,
  public = list(
    log_likelihood = function(data, params) {
      # Return S x N matrix of log-likelihoods
    },
    log_likelihood_gradient = function(data, params) {
      # Return list of S x N x K arrays (gradient per param)
    },
    log_likelihood_hessian_diag = function(data, params) {
      # Return list of S x N x K arrays (Hessian diagonal per param)
    }
  )
)
```

### Built-in likelihoods

- `LogisticRegressionLikelihood` -- binary classification
- `PoissonRegressionLikelihood` -- count data
- `LinearRegressionLikelihood` -- Gaussian regression

## Jacobian Correction

All small-step transformations compute the O(h) Jacobian correction term:

$$\log|J| \approx \log|1 + h \cdot \text{div}(Q)|$$

where $\text{div}(Q) = \text{tr}(\partial Q / \partial \theta)$ is the divergence of the
vector field.

- **LikelihoodDescent**: analytical divergence via Hessian diagonal
- **PMM2**: analytical divergence from the scaling ratios
- **KL, NKL**: numerical divergence via `numDeriv::jacobian`
- **MM1**: exact Jacobian = 1 (pure shift)
- **MM2**: exact Jacobian from affine scaling

## Controlling the Transformation Search

```{r control, eval=FALSE}
# Only run specific transformations
results <- sampler$adaptive_is_loo(
  data, params,
  transformations = c("identity", "kl", "nkl"),
  rhos = c(0.01, 0.1, 0.5, 1.0, 2.0),  # custom step sizes
  khat_threshold = 0.5,                  # stricter threshold
  try_all_transformations = TRUE         # don't skip adapted points
)
```

## References

- Chang, J.C. et al. (2024). "Gradient-flow adaptive importance sampling for
  Bayesian leave-one-out cross-validation." arXiv:2402.08151v2.
- Vehtari, A. et al. (2017). "Practical Bayesian model evaluation using
  leave-one-out cross-validation and WAIC." Statistics and Computing, 27(5).
